Summary of cell type ontologies in reference files

### 2024-12-17 analyses/cell-type-consensus/exploratory-notebooks/01-reference-exploration.Rmd
Apply suggestions from code review
Co-authored-by: Jaclyn Taroni <19534205+jaclyn-taroni@users.noreply.github.com>

This section will look at identifying the latest common ancestor (LCA) between all possible combinations of terms from PanglaoDB (used for assigning cell types with `CellAssign`) and the `BlueprintEncodeData` reference from `celldex` (used for assigning cell types with `SingleR`). 
The LCA refers to the latest term in the cell ontology heirarchy that is common between two terms. 
I will use the [`ontoProc::findCommonAncestors()` function](https://rdrr.io/bioc/ontoProc/man/findCommonAncestors.html) to get the LCA for each combination.

@@

```{r}
# get a data frame with all combinations of panglao and blueprint terms
# one row for each combination 
all_ref_df <- expand.grid(panglao_df$panglao_ontology,
                          blueprint_df$blueprint_ontology) |> 
  dplyr::rename(
    panglao_ontology = "Var1",
    blueprint_ontology = "Var2"
  ) |> 
  # add in the human readable values for each ontology term
  dplyr::left_join(blueprint_df, by = "blueprint_ontology") |> 
  dplyr::left_join(panglao_df, by = "panglao_ontology") |> 
  tidyr::drop_na() |> 
  dplyr::rowwise() |> 
  dplyr::mutate(
    # least common shared ancestor
    lca = list(rownames(ontoProc::findCommonAncestors(blueprint_ontology, panglao_ontology, g = g)))
  )

@@

Alternatively, rather than eliminate terms that are too broad we could look at the similarity index for individual matches and decide on a case by case basis if those should be allowed. 
Although I still think having a term that is too braod, even if it's a good match, is not super informative.

>>>>

This section will look at identifying the latest common ancestor (LCA) between all possible combinations of terms from PanglaoDB (used for assigning cell types with `CellAssign`) and the `BlueprintEncodeData` reference from `celldex` (used for assigning cell types with `SingleR`). 
The LCA refers to the latest term in the cell ontology hierarchy that is common between two terms. 
I will use the [`ontoProc::findCommonAncestors()` function](https://rdrr.io/bioc/ontoProc/man/findCommonAncestors.html) to get the LCA for each combination.

@@

```{r}
# get a data frame with all combinations of panglao and blueprint terms
# one row for each combination 
all_ref_df <- expand.grid(panglao_df$panglao_ontology,
                          blueprint_df$blueprint_ontology) |> 
  dplyr::rename(
    panglao_ontology = "Var1",
    blueprint_ontology = "Var2"
  ) |> 
  # add in the human readable values for each ontology term
  dplyr::left_join(blueprint_df, by = "blueprint_ontology") |> 
  dplyr::left_join(panglao_df, by = "panglao_ontology") |> 
  tidyr::drop_na() |> 
  dplyr::rowwise() |> 
  dplyr::mutate(
    # least common shared ancestor
    lca = list(rownames(ontoProc::findCommonAncestors(blueprint_ontology, panglao_ontology, g = cl_graph)))
  )

@@

Alternatively, rather than eliminate terms that are too broad we could look at the similarity index for individual matches and decide on a case by case basis if those should be allowed. 
Although I still think having a term that is too broad, even if it's a good match, is not super informative.
